Social Visual Analytics: DevOps for Data Scientists

### 2015-03-20 discussion.tex
working

\subsection{Design Considerations}

@@

{\bf Group management, size and diversity.} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. Rcloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

@@

{\bf Multiple language support.}
We did not forsee the demand to relax the assumption that programming
is done mainly in R. In retrospect, it is clear that many data science
problems are solved with cross-language approaches (for example, scikit-learn
in Python, weka in Java).\stephen{Isn't this fixed in some way now?}

{\bf Versioning.}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not explicitly
separate the development and deployment environments. More than that,
ever version of every prototype is shared by default. Although this encourages
collaboration on work in progress it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can unintentionally affect production websites. Such errors
are difficult to completely avoid, but the power of convenient sharing is often
worth it (as proven true with distributed systems in general).
Still, effective control over versioning when sharing problem needs more
attention if Rcloud would be scaled up to a wider number of users lacking
informal channels to coordinate work.
Versioning needs to be managed in several places: in scripts;
in the R environment such as the installed libraries and packages;
and in the external environment such as the operating system and
protocols spoken by remote services. At one end of the scale, we have
full control over the versioning of scripts via git, along with conventions
to name stable or working versions of scripts, and versions known to work together.
The R environment itself is under the control of its package manager
that has rules to ensure reasonable consistency.
It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a careful approach to system upgrades.

>>>>

\subsection{Reflection on Design Considerations}

@@

{\bf Group management, size and diversity.} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. RCloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

@@

{\bf Multiple language support.}
We did not foresee the demand to relax the assumption that programming
is done mainly in R. In retrospect, it is clear that many data science
problems are solved with cross-language approaches (for example, scikit-learn
in Python, weka in Java).\stephen{Isn't this fixed in some way now?}

{\bf Versioning.}
Some important aspects of the environment are more difficult to manage
in RCloud than in conventional systems. RCloud does not explicitly
separate the development and deployment environments. More than that,
ever version of every prototype is shared by default. Although this encourages
collaboration on work in progress it also quickly exposes errors such as
misconfiguration, mismatches between versions of packages, and programming
errors that can unintentionally affect production websites. Such errors
are difficult to completely avoid, but the power of convenient sharing is often
worth it (as proven true with distributed systems in general).
Still, effective control over versioning when sharing problem needs more
attention if RCloud would be scaled up to a wider number of users lacking
informal channels to coordinate work.
Versioning needs to be managed in several places: in scripts;
in the R environment such as the installed libraries and packages;
and in the external environment such as the operating system and
protocols spoken by remote services. At one end of the scale, we have
full control over the versioning of scripts via git, along with conventions
to name stable or working versions of scripts, and versions known to work together.
The R environment itself is under the control of its package manager
that has rules to ensure reasonable consistency.
It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a careful approach to system upgrades.

### 2015-03-20 introduction.tex
yes
working

\begin{addmargin}[1em]{1em}  % left and right margin change
\small
\stephen{Can we get away with \verb"\smaller"? Ha ha.}
\input{quotes}
\end{addmargin}

@@

Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
and sharing data from scientitic lab experiments\cite{Rubacha:ELN:2011}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.

>>>>

\begin{addmargin}[1em]{1em}  % left and right margin change
\small
\input{quotes}
\end{addmargin}

@@

Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:ELN:2011}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.

### 2015-03-20 relatedwork.tex
working

Kandel et al. argue in their review of data wrangling work that data
cleaning, wrangling and transformation is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. Although RCloud
does not by itself include modules specific to data cleaning and
wrangling, we note from practical experience that these cleaning
scripts themselves tend to change over time. RCloud addresses this
issue by providing easy publishing of data-cleaning notebooks as web
services, which reduces the total data-cleaning effort across an
organization.

Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. They also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is a difficult and important problem, and that sharing
and versioning of data sources and artifacts is hindered by current
technology in practice. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their work points to the opportunity for better technology
to support collaboration and sharing by data analysis teams.

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}.
Notebooks, and the integrated version control system for them, described in
Section~\ref{sec:notebooks}, address modularity and granularity, and
artifact histories.
\emph{Starring}, the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses cost of
integration, content export, presentation and view sharing.

Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
of social media with visualization and data publishing. We build on this
work by defining a rich interface for collaboration about code and for
operating on metadata.

@@

There has been noteworthy work on specific techniques such as
social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
or social development or analysis processes.
Similarly, there are computational methods to support high
performance execution in incremental code development
environments \cite{Guo:2010:TPI}.
Our goal is to define an environment in which many such
techniques may be integrated and made available to a broad community.

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer match for analysts accustomed to creating and
executing R and Python code, while retaining many attractive
properties like automatic versioning and management.

@@

Our work has been inspired by and benefited from other efforts
to improve data analysis environments and processes,
including RStudio \cite{RStudio:2013:SWA},
the R packages Markdown \cite{Allaire:2014:MMR},
knitr \cite{Xie:2013:DDW}
and Shiny \cite{RStudio:2013:SWA},
and iPython notebooks \cite{Perez:2007:IAS}
to name a few. RStudio aims at providing an integrated development environment
for R, with support for publishing code in packages. Markdown,
knitR and Shiny provide R with sophisticated reporting capabilities, including
interactive web interfaces. IPython \cite{Perez:2007:IAS}
shares many of our goals, such as providing a comprehensive environment
for analysis and programming, and shareable documents on the web.

One overall goal is to reduce the gap between implementers and deployers
of technology in visual analytics. The fusion of development with production
operations in software release management (``DevOps'' \cite{Httermann:2012:DD})
is a trend in web services and related fields. By making it possible
for data scientists to expend a little more effort in the process of
creating experiments, it may be possible to eliminate the need for
programming teams to recreate their results in production with
a concomitant penalty in time and expense.

>>>>

Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. Kandel et al also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is difficult and important, and that sharing
and versioning of data sources and artifacts is hindered by current
technology. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their work points to the opportunity for better technology
to support collaboration and sharing in data analysis teams.

An earlier study by Kandel et al argues that data wrangling
(cleaning, parsing and transformation)is a major part of exploratory
analysis and visualization~\cite{Kandel:2011:RDI}. We view this
as attacking a different semantic level than ours, but points
to the need for an environment that enables better sharing
of the knowledge, tools and processes to do this. Anecdotally
we find much frustration among practitioners that this knowledge
is difficult to find and often not recorded or provided in a reusable
form even within one organization.

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}.
RCloud notebooks, and the integrated version control system for them,
described in Section~\ref{sec:notebooks}, address modularity and granularity,
and artifact histories.
\emph{Starring}, the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses the cost of
integration, content export, presentation and view sharing.

Manyeyes \cite{Viegas:2007:MAS} was a landmark system for the integration
of social media with visualization and data publishing. We build on this
work by defining a rich interface for collaboration about code, and for
operating on metadata.

@@

There has been noteworthy work on specific techniques such as
social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
or social development or analysis processes.
Similarly, there are computational methods to support high
performance execution in incremental code development
environments \cite{Guo:2010:TPI}.
Our present goal is to define an environment in which many such
techniques may be integrated and made available to a broad community.

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer match for analysts accustomed to creating and
executing R and Python code, while retaining attractive
properties like automatic versioning and management.

@@

Our work has been inspired by and benefited from other efforts
to improve data analysis environments and processes,
including RStudio \cite{RStudio:2013:SWA},
the R packages Markdown \cite{Allaire:2014:MMR},
knitr \cite{Xie:2013:DDW}
and Shiny \cite{RStudio:2013:SWA},
and iPython notebooks \cite{Perez:2007:IAS}
to name a few. RStudio aims at providing an integrated development environment
for R, with support for publishing code in packages. Markdown,
knitR and Shiny augment R with sophisticated reporting capabilities, including
interactive web interfaces. IPython \cite{Perez:2007:IAS}
shares many of our goals, such as providing a comprehensive environment
for analysis and programming, and share-able documents on the web.
\stephen{Do we now need to introduce one of those remarks where we
explain that we are not the same as IPython.}

One overall goal is to reduce the gap between implementers and deployers
of technology in visual analytics. The fusion of development with production
operations in software release management (``DevOps'' \cite{Httermann:2012:DD}
or ``continuous integration'' \cite{Fowler:2006:Continuous}) is a trend in web
services and related fields.
By making it convenient for data scientists to expend just a little more effort
in the process of creating experiments, we may be able to eliminate the need
for programming teams to recreate their work in production which has a high
penalty in time, expense and sometimes correctness.
